---
title: "Coordinate Reference Systems"
format: html
editor: visual
---

# Kaggle Geospatial certificate CRS course in R

```{r}
library(tidyverse)
library(sf)
library(here)
library(units)
```

```{r}
regions <- read_sf(here("data/ghana/ghana/Regions/Map_of_Regions_in_Ghana.shp"))
```

```{r}
st_crs(regions)
```
```{r}
# Create a DataFrame with health facilities in Ghana
facilities_df <- read_csv(here("data/ghana/ghana/health_facilities.csv"))
```

There are a lot of missing lat/long values in the CSV that we need to drop.
`geopandas` just ignores them, but we need to filter them out for `sf`

```{r}
facilities_df %>% 
    summarise(across(c("Longitude", "Latitude"), ~ sum(is.na(.))))
```

```{r}
# Convert to sf object
# CRS 4326 = WGS84 (standard GPS coordinates)
facilities <- facilities_df %>% 
    filter(!is.na(Longitude), !is.na(Latitude)) %>% 
    st_as_sf(
        coords = c("Longitude", "Latitude"), # order: x = lon, y = lat
        crs = 4326,
        remove = FALSE  # Keep original lat/lon columns
    )
```

## Plots

```{r}
#| fig-width: 8
#| fig-height: 8

ggplot() +
    geom_sf(data = regions,
            linetype = "dotted") +
    geom_sf(data = st_transform(facilities, 32630),
            color = "blue",
            size = 1) +
    theme_minimal()
```

st_transform using **proj4** string

```{r}
st_transform(regions, "+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs")
```

## geometry attributes

```{r}
st_coordinates(head(st_geometry(facilities)))[,"X"]
```

```{r}
regions <- regions %>% 
    mutate(AREA = set_units(st_area(st_geometry(regions)), km^2))
```


```{r}
print(str_glue("The area of Ghana is {sum(regions$AREA)} square kilometers"))
```

# Exercises

```{r}
birds_df <- read_csv(here("data/purple_martin.csv"))
```

```{r}
colnames(birds_df) <- gsub(
    "\\.",
    "_",
    make.names(colnames(birds_df))
    )
unique_birds <- n_distinct(birds_df$tag_local_identifier)
print(str_glue("There are {unique_birds} different birds in the dataset."))
```
```{r}
birds <- birds_df %>% 
    st_as_sf(
        coords = c("location_long", "location_lat"),
        crs = 4326,
        remove = FALSE
    )
```

## Plot data

```{r}
world <- read_sf(here("data/tmap_World/tmap_World.shp"))
```

```{r}
americas <- world %>% 
    filter(continent %in% c("North America", "South America"))
```


```{r}
#| fig-width: 10
#| fig-height: 6

ggplot() +
    geom_sf(data = americas, fill = NA, color = "grey") +
    geom_sf(data = birds, size = 1, color = "blue") +
    theme_minimal()
```
```{r}
birds
```

### creates linestrings for the paths of each bird and the starting and ending point for each bird's journey

```{r}
path_gdf <- birds %>% 
    arrange(tag_local_identifier, timestamp) %>% # ensures points are ordered by timestamp
    summarize(
        .by = tag_local_identifier,
        # converts the MULTIPOINT obj from st_combine into a LINESTRING
        geometry = st_combine(geometry) %>% st_cast("LINESTRING")
    ) %>% 
    st_as_sf()
    
head(path_gdf)
```

```{r}
start_gdf <- birds %>% 
    arrange(tag_local_identifier, timestamp) %>% # ensures points are ordered by timestamp
    summarize(
        .by = tag_local_identifier,
        geometry = geometry[1]
    ) %>% 
    st_as_sf()

head(start_gdf$geometry %>% st_coordinates())
``` 


```{r}
end_gdf <- birds %>% 
    arrange(tag_local_identifier, timestamp) %>% # ensures points are ordered by timestamp
    summarize(
        .by = tag_local_identifier,
        geometry = geometry[length(geometry)]
    ) %>% 
    st_as_sf()

head(end_gdf$geometry %>% st_coordinates())
```
```{r}
head(end_gdf)
```

```{r}
#| fig-width: 10
#| fig-height: 6

ggplot() +
    geom_sf(data = americas, fill = NA, color = "grey") +
    geom_sf(data = start_gdf, aes(color = factor(tag_local_identifier)), size = 2) +
    geom_sf(data = path_gdf, aes(color = factor(tag_local_identifier)), linewidth = 1) +
    geom_sf(data = end_gdf, aes(color = factor(tag_local_identifier)), size = 2) +
    scale_color_manual(values = 
                           c("red",
                             'orange',
                             'gold',
                             'green',
                             'blue',
                             'purple',
                             'salmon',
                             'cyan',
                             'burlywood',
                             'cadetblue',
                             'maroon'
                             )
                       ) +
    theme_minimal()
```

## Protected areas calculations

```{r}
protected_filepath <- here("data/SAPA_Aug2019-shapefile/SAPA_Aug2019-shapefile/SAPA_Aug2019-shapefile-polygons.shp")
protected_areas <- read_sf(protected_filepath)
```

### shapefile is big 

rewrite as geojson for compression with xz

```{r}
st_write(protected_areas,
         here("data/SAPA_Aug2019-shapefile/SAPA_Aug2019-shapefile/SAPA_Aug2019-shapefile-polygons.geojson"))
```
FlatGeobuf with zstd compresses about as well as GeoJson, but the unpacked file is much smaller

```{r}
st_write(protected_areas,
         here("data/SAPA_Aug2019-shapefile/SAPA_Aug2019-shapefile/SAPA_Aug2019-flatgeobuf-polygons.fgb"))

```

```{r}
head(protected_areas)
```
```{r}
south_america <- americas %>% 
    filter(continent == "South America")
```

```{r}
#| fig-width: 10
#| fig-height: 6

ggplot() +
    geom_sf(data = south_america, fill = NA, color = "grey") +
    geom_sf(data = protected_areas, fill = "blue", color = "blue")

```

```{r}
P_Area <- sum(protected_areas$REP_AREA - protected_areas$REP_M_AREA)
print(str_glue("South America has {P_Area} square kilometers of protected areas."))
```

```{r}
head(south_america)
```
```{r}
st_crs(south_america)
```

```{r}
south_america <- st_transform(south_america, 3035)
```


```{r}
totalArea <- 
    set_units(
        sum(
            st_area(
                st_geometry(
                    st_transform(
                        south_america, 
                        3035
                    )
                )
            )
        ),
        km^2
    )

print(totalArea)

```

```{r}
# What percentage of South America is protected?
percentage_protected <- round((P_Area/totalArea) * 100, 2)
print(str_glue("Approximately {percentage_protected}% of South America is protected."))
```

```{r}
nom_protected_areas <- protected_areas %>% 
    filter(MARINE != 2)

birds_sa <- birds %>% 
    filter((st_coordinates(st_geometry(.))[,"Y"]) < 0)
```
```{r}
birds %>% filter((st_coordinates(st_geometry(.))[,"Y"]) < 0)
```

```{r}
#| fig-width: 10
#| fig-height: 10

ggplot() +
    geom_sf(data = south_america, fill = "white", color = "grey") +
    geom_sf(data = nom_protected_areas, fill = "skyblue", color = NA, alpha = 0.4) +
    geom_sf(data = birds_sa, color = 'red', alpha = 0.6, size = 2) +
    theme_minimal()
```

