---
title: "Geocoding and Geospatial joins"
author: "Jay Spiros"
date: "`r Sys.Date()`"
format: html
editor: source
---

```{r}
#| label: load-libs
#| message: false
#| warning: false
#| output: false

library(here)
library(tidyverse)
library(janitor)
library(sf)
library(leaflet)
library(tidygeocoder)
```

# Introduction

In this tutorial, you'll learn about two common manipulations for geospatial data: geocoding and table joins.

```{r}
location <- tibble(place = "Fenway Park") %>% 
    geocode(place, method = "osm", lat = latitude, long = longitude, full_results = TRUE)
```

```{r}
location$display_name
```

```{r}
location %>% 
    select(latitude, longitude)
```

```{r}
universities = read_csv(here("data/top_universities.csv"))
head(universities)
```

```{r}
nrow(universities)
```

```{r}
universities_coded <- universities %>% 
    geocode(Name, method = "osm", lat = Latitude, long = Longitude)
```

```{r}
# perc <- (1 - sum(is.na(universities_coded$Latitude)) / nrow(universities_coded)) * 100
perc <- sum(!is.na(universities_coded$Latitude)) / nrow(universities_coded) * 100
print(str_glue("{perc}% of universities geocoded"))
```

```{r}
universities_cleaned <- drop_na(universities_coded)
```

```{r}
perc_cleaned <- (1 - sum(is.na(universities_cleaned$Latitude)) / nrow(universities_cleaned)) * 100
print(str_glue("{perc_cleaned}% of universities geocoded"))
```

```{r}
universities <- universities_cleaned %>% 
    st_as_sf(
        coords = c("Longitude", "Latitude"),
        crs = 4326,
        remove = FALSE
    )
```

```{r}
head(universities)
```

```{r}
leaflet(universities) %>% 
    addTiles() %>% 
    setView(lng = 15, lat = 54, zoom = 2) %>% 
    addMarkers(lng = ~Longitude, 
               lat = ~Latitude,
               clusterOptions = markerClusterOptions(),
               popup = ~Name
    )
```

# Spatial joins

Use `st_join` In particular the join = argument is important

Docs: <https://r-spatial.github.io/sf/reference/st_join.html>

# Exercises

```{r}
#| message: false

starbucks <-  read_csv(here("data/starbucks_locations.csv"))
starbucks <-  clean_names(starbucks)
```

```{r}
starbucks %>% summarise(across(everything(), ~ sum(is.na(.))))
```

```{r}
rows_missing <- is.na(starbucks$longitude) | is.na(starbucks$latitude)

starbucks[rows_missing, c("longitude", "latitude")] <-
    geocode(starbucks[rows_missing, "address"], address, method = "osm")[ ,c("long", "lat")]
```

```{r}
starbucks[starbucks$city == "Berkeley", ]
```

### these accomplish the same thing (tidyverse, from Claude)

```{r}
df <- df |>
  mutate(needs_geocode = is.na(longitude) | is.na(latitude)) |>
  group_by(needs_geocode) |>
  group_modify(~ {
    if (.y$needs_geocode) {
      geocode(.x, address, method = "osm") |>
        mutate(longitude = long, latitude = lat) |>
        select(-long, -lat)
    } else {
      .x
    }
  }) |>
  ungroup() |>
  select(-needs_geocode)
```

This splits the dataframe into two groups (needs geocoding / doesn't), geocodes only the group that needs it, then reassembles. It's more verbose but stays entirely within tidy pipelines.

A more concise tidyverse alternative using newer rows_update():

```{r}
geocoded <- df |>
  filter(is.na(longitude) | is.na(latitude)) |>
  select(address) |>
  geocode(address, method = "osm") |>
  rename(longitude = long, latitude = lat)

df <- df |>
  rows_update(geocoded, by = "address", unmatched = "ignore")
```

```{r}
leaflet(starbucks %>% filter(city == "Berkeley")) %>% 
    addTiles() %>% 
    setView(lng = -122.26, lat = 37.88, zoom = 13) %>% 
    addMarkers(lng = ~longitude, 
               lat = ~latitude,
               # clusterOptions = markerClusterOptions(),
               popup = ~store_name)
```


## Joins

```{r}
CA_counties <- 
    read_sf(here("data/CA_county_boundaries/CA_county_boundaries/CA_county_boundaries.shp"))

```

```{r}
#| message: false

# this CSV has leading zeros in GEOID leading readr to think it's a char column
CA_median_age <- read_csv(here("data/CA_county_median_age.csv"), col_type = "d")
CA_pop <- read_csv(here("data/CA_county_population.csv"))
CA_high_earners <- read_csv(here("data/CA_county_high_earners.csv"))

```

```{r}
CA_dataframes <- list(CA_counties, CA_pop, CA_median_age, CA_high_earners)
CA_stats <- reduce(CA_dataframes, left_join, by = "GEOID")
```

```{r}
CA_stats <- 
    CA_stats %>% 
    mutate(density = population / area_sqkm)
```

```{r}
head(CA_stats)
```

### 4) Which counties look promising?

Collapsing all of the information into a single GeoDataFrame also makes it much easier to select counties that meet specific criteria.

Use the next code cell to create a GeoDataFrame `sel_counties` that contains a subset of the rows (and all of the columns) from the `CA_stats` GeoDataFrame. In particular, you should select counties where:

- there are at least 100,000 households making \$150,000 per year,

- the median age is less than 38.5, and

- the density of inhabitants is at least 285 (per square kilometer).

Additionally, selected counties should satisfy at least one of the following criteria:

- there are at least 500,000 households making \$150,000 per year,

- the median age is less than 35.5, or

- the density of inhabitants is at least 1400 (per square kilometer).

```{r}
sel_counties <- 
    CA_stats %>% 
    filter(high_earners >= 100000,
               median_age < 38.5,
               density >= 285
               ) %>% 
    filter(when_any(
        high_earners >= 500000,
        median_age < 35.5,
        density >= 1400
    ))
    
```

```{r}
sel_counties
```

### transform starbucks dataframe to sf

```{r}
starbucks_gdf <- 
    st_as_sf(starbucks,
             coords = c("longitude", "latitude"),
             crs = 4326,
             remove = FALSE
             )
```

```{r}
starbucks_gdf
```
```{r}
ggplot() +
    geom_sf(data = sel_counties, fill = NA) +
    #geom_sf(data = starbucks_gdf, color = "green") +
    theme_minimal()
```

```{r}
possible_stores <- st_join(sel_counties, starbucks_gdf) 
num_stores <- nrow(possible_stores)
```

```{r}
print(num_stores)
print(nrow(starbucks_gdf))
```
```{r}
head(possible_stores)
```
```{r}
# if we don't do this leaflet will redraw the county polygon once for each row in the df, which is over 1000.
# this creates a sub frame with the county polygons in there only once per county.
# Could also go back the CA_counties dataframe and filter for the counties we want
# but if the selected counties change we'd have to rewrite the filter
county_borders <- possible_stores %>% 
    group_by(name) %>% 
    slice_head(n = 1)
```


```{r}
leaflet(possible_stores) %>% 
    addTiles() %>% 
    setView(lng = -120, lat = 37, zoom = 6) %>%
    addMarkers(lng = ~longitude, 
               lat = ~latitude,
               clusterOptions = markerClusterOptions(),
               popup = ~store_name) %>% 
    addPolygons(data = county_borders,
                color = "black",
                weight = 1,
                opacity = 0.8,
                #fillColor = "none",
                #fillColor = "lightgreen",
                fillOpacity = 0,
                # highlightOptions = highlightOptions(
                #     weight = 1,
                #     opacity = 1,
                #     fillOpacity = 0.7,
                #     bringToFront = TRUE,
                #     sendToBack = TRUE
                # ),
                label = ~name,
                labelOptions = labelOptions(
                    style = list("font-weight" = "normal", padding = "3px 8px"),
                    textsize = "15px",
                    direction = "auto"
                )
    )
    
```

